
% Thesis Abstract -----------------------------------------------------


%\begin{abstractslong}    %uncommenting this line, gives a different abstract heading
\begin{abstracts}        %this creates the heading for the abstract page
A new quantum Monte Carlo method based on stochastically sampling the thermal density matrix at finite temperatures has been developed. The method is an adaptation of the recently developed full configuration interaction quantum Monte Carlo\cite{Booth2009} (FCIQMC) method and it aims to overcome two of FCIQMCs main restrictions by allowing the calculation of any quantum mechanical observable at finite temperatures. During this report the density matrix quantum Monte Carlo (DMQMC) algorithm is formulated by observing an analogy with FCIQMC. It is then tested on the spin-$1/2$ antiferromagnetic Heisenberg model on a bipartite lattice.

The DMQMC method produces results that agree with the exact finite temperature FCI energy and exact ground-state staggered magnetisation for the $4\times4$ lattice. For the larger $6\times6$ lattice the standard DMQMC algorithm fails.  At this point an importance sampling method is introduced, yielding results comparable to accurate Green's function quantum Monte Carlo (GFQMC) estimates. The importance sampling method also enabled simulation of the $4\times4\times4$ lattice, producing results for the ground-state energy that matched those obtained by FCIQMC.

Finally an avenue into quantum information theory is explored and some key results for the entanglement between two qubits on a spin-$1/2$ antiferromagnetic Heisenberg ring are reproduced.  Additionally, both how the entanglement between qubits changes in a uniformly applied magnetic field and with increasing separation between the two qubits are investigated. 

\end{abstracts}
%\end{abstractlongs}


% ----------------------------------------------------------------------


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../thesis"
%%% End: 
